Speeding up decimal multiplication

### 2020-11-19 paper/paper.tex
Update

\begin{abstract}
    Decimal multiplication is the task of multiplying two numbers in base $10^N.$
    Specifically, we focus on the number-theoretic transform (NTT) family of algorithms.
    Using only portable techniques, we achieve a 3x---5x speedup over the \textbf{mpdecimal} library.
    In this paper we describe our implementation and discuss further possible optimizations.
    Our contributions are the following:
    (1) the ``linearity trick'' for using Montgomery reduction with NTT without separate steps of conversion into/out of Montgomery form;
    (2) a simple cache-oblivious algorithm for in-place $2n \times n$ or $n \times 2n$ matrix transposition, the need for which arises in the ``six-step algorithm'' variation of the matrix Fourier algorithm, and which does not seem to be widely known.
    Notable findings include:
    (1) use of two prime moduli instead of three makes sense even considering the worst case of increasing the size of the input, and makes for simpler answer recovery;
    (2) the ``six-step algorithm'', supposedly fast for large transforms due to cache-friendliness, is of little use on modern systems for transforms of sizes up to $\approx 2^{27}.$
\end{abstract}

@@

Unfortunately, this method is very slow. % TODO

@@

We found that on a modern x86-64 system, reduction with two rounds of $r$ is slower by
$\approx$ 15\% compared to Montgomery reduction;
the code can be found in the \verb!bench-solinas! subdirectory of our repository.

@@

IDFT of length $N = RC$ can be computed as follows:
\begin{enumerate}
    \item $[C \times R],$ transpose the matrix.
    \item $[R \times C],$ perform (length $C$) IDFT on each row with $\xi = (\xi_N)^R.$
    \item $[R \times C],$ multiply each element indexed $\langle i,j \rangle$ by $(\xi_N)^{-ij}.$
    \item $[R \times C],$ perform (length $R$) IDFT on each column with $\xi = (\xi_N)^C.$
\end{enumerate}

@@

IDFT of length $N = RC$ can be computed as follows:
\begin{enumerate}
    \item $[C \times R],$ transpose the matrix.
    \item $[R \times C],$ perform (length $C$) IDFT on each row with $\xi = (\xi_N)^R.$
    \item $[R \times C],$ multiply each element indexed $\langle i,j \rangle$ by $(\xi_N)^{-ij}.$
    \item $[R \times C],$ transpose the matrix.
    \item $[C \times R],$ perform (length $R$) IDFT on each row with $\xi = (\xi_N)^C.$
    \item $[C \times R],$ transpose the matrix.
\end{enumerate}

@@

In its general form, in-place transposition of a non-square matrix is quite hard.
The classical study begins with defining the function $\mathcal{P}(a)$ such that,
when transposing a matrix with $N_1$ rows and $N_2$ columns in row-major order,
the element with index $a$ is sent to index $\mathcal{P}(a).$
There is, then, a result which says that
\begin{equation}
\mathcal{P}(a) = \begin{cases}
    N_1 N_2 - 1                 & \text{if } a = N_1 N_2 - 1; \\
    N_1 a \mymod (N_1 N_2 - 1)  & \text{otherwise.}
\end{cases}
\end{equation}
The number of fixed points of $\mathcal{P}$ is exactly
$$
1 + \gcd(N_1-1, N_1-1).
$$
The number of cycles of length $k>1$ of $\mathcal{P}$ is given by
$$
\frac{1}{k} \sum\limits_{d | k} \widetilde{\mu}(k/d) \gcd(N_1^d - 1, N_1 N_2 - 1),
$$
where $\widetilde{\mu}$ is the Möbius function and the summation is performed over all divisors of
$k.$

Most algorithms for in-place transposition are essentially of ``follow-the-cycles'' kind.
This means that they iterate over all the cycles, and for each cycle $C,$
they shift the elements of $C$ cyclically. There are, then, various approaches for
locating the cycles, identifying the first element of a cycle, and tracking which cycles were
already visited.
If we limit ourselves to using only $O(N_1+N_2)$ additional memory,
even more complex algorithms are genreally needed;
\cite{Fich95permutingin}
presents an algorithm with worst-case complexity of $O(N_1 N_2 \log(N_1 N_2)).$

@@

\subsection{Algorithm for matrix transposition}

@@

Note it uses the routine $\mathrm{TransposeSquareSubMatrix}$.
It does not have to be implemented exactly in a way we demonstrated above;
it just needs to have the same semantics.
Note that it can be implemented in a cache-friendly way by dividing the square submatrix into
tiles, perhaps recursively, transposing the elements inside tiles, and then transposing the tiles
themselves; see~\cite{cache_square_mat_trans} for details.

@@

@@

@@

Note this is far from being a tight bound in numerical sense:
in practice, the value of
$\frac{L}{(B-1)M}=B-1$ is going to be much larger than 1, meaning that the exact upper bound on
$c_i + \sigma_i$ is going to be closer to $\mu^2 / 4$ than to $\mu^2 / 2;$ but this is still a
tight bound in the sense that maximum possible $c_i + \sigma_i$ would not into one machine word.
This bound also has the advantage that it works for any $B > 1.$

@@

>>>>

\begin{abstract}
    Decimal multiplication is the task of multiplying two numbers in base $10^N.$
    Specifically, we focus on the number-theoretic transform (NTT) family of algorithms.
    Using only portable techniques, we achieve a 3x---5x speedup over the \textbf{mpdecimal} library.
    In this paper we describe our implementation and discuss further possible optimizations.
    Our contributions are the following:
    (1) the ``linearity trick'' for using Montgomery reduction with NTT without separate steps of conversion into/out of Montgomery form;
    (2) a simple cache-efficient algorithm for in-place $2n \times n$ or $n \times 2n$ matrix transposition, the need for which arises in the ``six-step algorithm'' variation of the matrix Fourier algorithm, and which does not seem to be widely known.
    Notable findings include:
    (1) use of two prime moduli instead of three makes sense even considering the worst case of increasing the size of the input, and makes for simpler answer recovery;
    (2) the ``six-step algorithm'', supposedly fast for large transforms due to cache-friendliness, is of little use on modern systems for transforms of sizes up to $\approx 2^{27}.$
\end{abstract}

@@

Unfortunately, this method is very slow; in our benchmark, native division on x86-64
was 720\% slower than the Montgomery reduction. The code can be found in the
\verb!bench-modmul! subdirectory of our repository.

@@

We found that on a modern x86-64 system, reduction with two rounds of $r$ is slower by
$\approx$ 15\% compared to Montgomery reduction;
the code can be found in the \verb!bench-modmul! subdirectory of our repository.

@@

IDFT of length $N = RC$ can be computed as follows, not counting final multiplication by $N^{-1}$:
\begin{enumerate}
    \item $[C \times R],$ transpose the matrix.
    \item $[R \times C],$ perform (length $C$) IDFT on each row with $\xi = (\xi_N)^R.$
    \item $[R \times C],$ multiply each element indexed $\langle i,j \rangle$ by $(\xi_N)^{-ij}.$
    \item $[R \times C],$ perform (length $R$) IDFT on each column with $\xi = (\xi_N)^C.$
\end{enumerate}

@@

IDFT of length $N = RC$ can be computed as follows, not counting final multiplication by $N^{-1}$:
\begin{enumerate}
    \item $[C \times R],$ transpose the matrix.
    \item $[R \times C],$ perform (length $C$) IDFT on each row with $\xi = (\xi_N)^R.$
    \item $[R \times C],$ multiply each element indexed $\langle i,j \rangle$ by $(\xi_N)^{-ij}.$
    \item $[R \times C],$ transpose the matrix.
    \item $[C \times R],$ perform (length $R$) IDFT on each row with $\xi = (\xi_N)^C.$
    \item $[C \times R],$ transpose the matrix.
\end{enumerate}

@@

In its general form, in-place transposition of a non-square matrix is quite hard.
The classical study begins with defining the function $\mathcal{P}(a)$ such that,
when transposing a matrix with $N_1$ rows and $N_2$ columns in row-major order,
the element with index $a$ is sent to index $\mathcal{P}(a).$
There is, then, a result which says that
\begin{equation}
\mathcal{P}(a) = \begin{cases}
    N_1 N_2 - 1                 & \text{if } a = N_1 N_2 - 1; \\
    N_1 a \mymod (N_1 N_2 - 1)  & \text{otherwise.}
\end{cases}
\end{equation}
The number of fixed points of $\mathcal{P}$ is exactly
$$
1 + \gcd(N_1-1, N_2-1).
$$
The number of cycles of length $k>1$ of $\mathcal{P}$ is given by
$$
\frac{1}{k} \sum\limits_{d | k} \widetilde{\mu}(k/d) \gcd(N_1^d - 1, N_1 N_2 - 1),
$$
where $\widetilde{\mu}$ is the Möbius function and the summation is performed over all divisors of
$k.$

Most algorithms for in-place transposition are essentially of ``follow-the-cycles'' kind.
This means that they iterate over all the cycles, and for each cycle,
they shift its elements cyclically. There are, then, various approaches for
locating the cycles, identifying the first element of a cycle, and tracking which cycles were
already visited.
If we limit ourselves to using only $O(N_1+N_2)$ additional memory,
even more complex algorithms are genreally needed;
\cite{Fich95permutingin}
presents an algorithm with worst-case complexity of $O(N_1 N_2 \log(N_1 N_2)).$

@@

\subsection{Algorithm for matrix transposition}
\label{section:mat_trans_algo}

@@

Note it uses the routine $\mathrm{TransposeSquareSubMatrix}$.
It does not have to be implemented exactly in a way we demonstrated above;
it just needs to have the same semantics.
It can be implemented in a cache-friendly way by dividing the square submatrix into
tiles, perhaps recursively, transposing the elements inside tiles, and then swapping the tiles
themselves; see~\cite{cache_square_mat_trans} for details.

@@

With $N=3 \cdot 2^k,$ we interpret the sequence as a matrix with 3 rows and $2^k$ columns.

Observe that steps 1 and 2 of the direct transform, namely
\begin{itemize}
\item length-3 DFT on each column,
\item multiplication by $(\xi_N)^{ij}$),
\end{itemize}
as well as steps 3 and 4 of the inverse transform, namely
\begin{itemize}
\item multiplication by $(\xi_N)^{-ij},$
\item length-3 IDFT on each column,
\end{itemize}
can be merged (performed in one pass over the columns).
In fact, they should be merged, as it produces less loads/stores and leads to better
cache utilization.

If, additionally, we need to multiply each matrix element by some factor $\Phi,$
then we can meld this operation into the column operations as well; and, as it turns out,
this can save us a few multiplications per column.
Note there is no difference at which step to multiply the matrix by a scalar $\Phi,$
because all the column operations involved are linear.

Note that we do need to multiply everything by $N^{-1}$ after performing the inverse transform
(or by another factor if the linearity trick is used; see section~\ref{section:trick}).
For the direct transform, we only need to ``multiply'' (perform $\redc$ with $R^2$)
if the linearity trick is not used.

Suppose we need to perform length-3 DFT on each column, then multiply each matrix element indexed
$\langle i,j \rangle$
by $(\xi_N)^{ij};$ and also multiply everything by a factor $\Phi.$
Define $\Gamma = \xi_N,$ $\Lambda = (\xi_N)^{N/3}.$
This operation on a signle column with index $j,$
$\langle x, y, z \rangle \mapsto \langle x', y', z' \rangle,$ can be expresed as:

\begin{subequations}
\begin{align}
    x' &= \Phi (x + y + z); \\
    y' &= \widetilde{\Phi}_1 (x + y \Lambda   + z \Lambda^2); \\
    z' &= \widetilde{\Phi}_2 (x + y \Lambda^2 + z \Lambda),
\end{align}
\end{subequations}
where
\begin{subequations}
\begin{align}
    \widetilde{\Phi}_1 &= \Phi \Gamma^j; \\
    \widetilde{\Phi}_2 &= \Phi \Gamma^{2j}.
\end{align}
\end{subequations}

Note that $\widetilde{\Phi}_1$ and $\widetilde{\Phi}_2$ can be computed incrementally,
with two variables initially having the value of $\Phi$ and get multiplied by
$\Gamma$ and $\Gamma^2,$ correspondingly, after each iteration of the loop as we operate on columns
$0, \ldots, 2^k-1.$ This is an improvement over a separate multiply-by-$\Phi$ step:
we have 9 multiplications per column instead of 11.
Note that if we do not need to multiply by $\Phi,$ we have 8 multiplications per column.

Suppose now we need to multiply each matrix element indexed $\langle i,j \rangle$
by $(\xi_N)^{-ij},$ then perform length-3 IDFT on each column; and also multiply everything by a
factor $\Phi.$

Define $\Gamma = (\xi_N)^{-1},$ $\Lambda = (\xi_N)^{-N/3}.$
This operation on a signle column with index $j,$
$\langle x, y, z \rangle \mapsto \langle x', y', z' \rangle,$ can be expresed as:

\begin{subequations}
\begin{align}
    x' &= \widetilde{x} + \widetilde{y}           + \widetilde{z}; \\
    y' &= \widetilde{x} + \widetilde{y} \Lambda   + \widetilde{z} \Lambda^2; \\
    z' &= \widetilde{x} + \widetilde{y} \Lambda^2 + \widetilde{z} \Lambda,
\end{align}
\end{subequations}
where
\begin{subequations}
\begin{align}
    \widetilde{x} &= x \cdot \Phi; \\
    \widetilde{y} &= y \cdot \Phi \Gamma^j; \\
    \widetilde{z} &= z \cdot \Phi \Gamma^{2j}.
\end{align}
\end{subequations}

Again, the factors for $\widetilde{y}$ and $\widetilde{z},$
that is $\Phi \Gamma^j$ and $\Phi \Gamma^{2j},$ can be computed incrementally.
Similarly to the case of direct transform, this is an improvement over a
separate multiply-by-$\Phi$ step: 9 multiplications per column instead of 11.

@@

We need to factorize $N = 2^k$ into $N_1 \cdot N_2 = N$ so that $\max\{N_1, N_2\}$ is as small as
possible.
This is because we will perform transforms of lengths both $N_1$ and $N_2,$ and we want to achieve
good cache utilization.
This is done by choosing $N_1 = 2^{\lfloor k / 2 \rfloor}, N_2 = 2^{\lceil k / 2 \rceil}.$
Note that in case of even $k,$ the matrix is square; in case of odd $k,$
$N_2 = 2 \cdot N_1$ and the matrix can be transposed with the algorithm we described in
section~\ref{section:mat_trans_algo}.

Just like with the four-step algorithm, it is desirable to merge together the steps of
multiplication by $(\xi_N)^{\pm ij}$ and multiplication by a scalar $\Phi,$
and compute the factor $\Phi (\xi_N)^{\pm ij}$ for the current element at
$\langle i,j \rangle$ incrementally.
Compared to a separate multiply-by-$\Phi$ step, this approach
lowers the number of multiplications from $3N + o(N)$ to $2N + o(N).$

@@

Note this is far from being a tight bound in numerical sense:
in practice, the value of
$\frac{L}{(B-1)M}=B-1$ is going to be much larger than 1, meaning that the exact upper bound on
$c_i + \sigma_i$ is going to be closer to $\mu^2 / 4$ than to $\mu^2 / 2;$ but this is still a
tight bound in the sense that the maximum possible $c_i + \sigma_i$ would overflow one machine word.
This bound also has the advantage that it works for any $B > 1.$

@@

\subsection{Cache aliasing problems}
